---
title: "District Middle School: Data Analysis and Plan"
date: "2-12-2018"
output: 
  ioslides_presentation:
    css: jtc_styles.css
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      fig.width = 7,
                      fig.asp = 0.618,
                      fig.align = "center")


# read in the data, and organize it into a tidy format

library(tidyverse)
library(stringr)
library(modelr)
library(forcats)


source("jtc_theme.R")

demos <- read_csv("demos.csv")

test <- read_csv("test_scores.csv")

# actually some of the items in demos and others are related
# to discipline, so we should extract those
demos_tidy <- demos %>% 
  slice(1:7) %>% 
  gather(year, value, 2:5) %>% 
  rename(category = X1) %>% 
  mutate(value = str_extract(value, "[0-9]+"),
         pct = as.numeric(value) * .01) # extract the percentages

# add in a total students for each year
total <- demos_tidy %>% 
  filter(str_detect(category, "Total")) %>% 
  select(-category,
         -pct,
         total = value) %>% 
  mutate(total = as.numeric(total))

# and then multiply by percent to give the number in each category

demos_tidy <- demos_tidy %>% 
  filter(!str_detect(category, "Total")) %>% 
  left_join(total, by = "year") %>% 
  mutate(count = total * pct,
         category = str_replace(category, "% ", ""),
         category = recode(category, 
                           "Students as English Language Learners" = "ELLs",
                           "Students with Disabilities" = "IEP")) %>% 
  select(-value)

# now moving on to outcomes/test scores
out <- test %>% 
  gather(year, pct, 3:6) %>% 
  mutate(subj = str_extract(`Academic Years`, "reading|math"),
         pct = as.numeric(str_replace(pct, "%", "")) * .01) %>% 
  select(-`Academic Years`,
         grade = X2) %>% 
  unite(obs, grade, subj, sep = " ") %>% 
  mutate(obs = paste0(obs, " prof")) 

# we will add % with >95% attendance and pct with 0 suspensions (climate) 
# into the outcomes table, because it might be helpful to look at them together
climate <- demos %>% 
  slice(10:11) %>% 
  gather(year, pct, 2:5) %>% 
  mutate(obs = if_else(str_detect(X1, "Suspensions"), 
                       "0 suspensions", 
                       "95% attendance"),
         pct = as.numeric(str_replace(pct, "%", "")) * .01) %>% 
  select(-X1)

out <- bind_rows(out, climate)

# climate will remain separate because the unit of obervation is
# a referral, not a student
discipline <- demos %>% 
  slice(8:9) %>% 
  gather(year, value, 2:5) %>% 
  spread(X1, value) %>% 
  select(1,referrals = 2, pct_susp = 3) %>% 
  mutate(pct_susp = as.numeric(str_replace(pct_susp, "%", "")) * .01,
         suspensions = as.numeric(referrals) * pct_susp) %>% 
  select(-pct_susp) %>% 
  gather(type, count, 2:3) %>% 
  mutate(count = as.numeric(count))

```

## Finding 1

Student body size drecreased until most recent school year, while referrals and suspensions increased until most recent year, when each pattern reversed.

```{r}
demos_tidy %>% 
  filter(!category %in% c("IEP", "ELLs")) %>% 
  group_by(year) %>% 
  summarise(count = sum(count)) %>% 
  mutate(type = "total students") %>% 
  bind_rows(discipline) %>% 
  mutate(type = fct_relevel(type, "total students", 
                            "referrals",
                            "suspensions")) %>% 
  ggplot(aes(x = year, y = count, group = type, color = type)) + 
    geom_line(size = 2) +
    labs(title = str_wrap("Referrals follow opposite pattern of student body size",
                          35), 
         color = NULL,
         x = "school year") + 
    jtc
```

## Finding 2 

Substantial increase in the percent of White students. 

```{r}
demos_tidy %>% 
  filter(year %in% c("2013-2014", "2016-2017")) %>% 
  select(-pct, -total) %>% 
  spread(year, count) %>% 
  mutate(pct_change = (`2016-2017` - `2013-2014`) / `2013-2014`, 
         category = recode(category, Other = "other race"),
         type = if_else(str_detect(category, "ELLs|IEP"),
                                   "support services",
                                   "race"),
         category = str_wrap(category, 10), 
         category = fct_reorder(category, as.numeric(factor(type)))) %>% 
  ggplot(aes(category, pct_change)) +
    geom_bar(stat = "identity") + 
    scale_y_continuous(limits = c(-3, 3)) +
    labs(y = "percent change",
         x = "category", 
         title = "Change in demographic categories",
         subtitle = "SY13-14 to SY1617") + 
    jtc 

```

## Finding 3

Test scores decrease when the student body is more African-American, but increase with each other category. 

```{r}
test_scores <- out %>% 
  filter(str_detect(obs, "prof")) 

model <- lm(pct ~ obs, data = test_scores)

test_scores %>% 
  add_predictions(model) %>% 
  add_residuals(model) %>% 
  left_join(demos_tidy %>% rename(pct_demo = pct), by = "year") %>% 
  mutate(category = recode(category, Other = "other race"),
         type = if_else(str_detect(category, "ELLs|IEP"),
                                   "support services",
                                   "race"),
         category = fct_reorder(category, as.numeric(factor(type)))) %>% 
  ggplot(aes(pct_demo, resid)) +
    geom_jitter(aes(color = category), alpha = 0.5, show.legend = FALSE) + 
    geom_smooth(aes(color = category), show.legend = FALSE, size = 1.5,
                method = "glm", se = FALSE) + 
    facet_wrap(~category, scales = "free_x") +
    labs(title = "Test scores change with demographics", 
         x = "percent category",
         y = "residuals - test scores") + 
    jtc +
    theme(text = element_text(size = 11))
```

## Further Exploration

#### Evidence suggests that positive changes in recent year may reflect more about changes in student body make-up than school culutre

- Additional data exploration is needed to understand how school and neighborhood context combine to increase student achievement
    - disaggregate test score and discipline information by race
- 

## Strategic plan
- 

## Turnaround Principle 6: Safe school environment

#### Rationale
- High rates of discipline referrals and suspension
- Decrease in most recent year may or may not reflection positive change

#### Work with principal to:
- Understand discipline procedures and systems
- Explore disagreggated discipline data and trends over time

## Turnaround Principle 6: Safe school environment

- Conduct surveys of staff, students, and families
- Implement school-wide positive behavioral supports
- Encourage restorative practices

## Parent tip

### Academic Parent-Teacher Teams (APTT)

Adapated from [this](https://vimeo.com/161258860) resource for middle school students and focusing on safe school environment

- Initial focus on behavioral supports transitioning to specific academic standards
- Review data anonymized data about student behavior and achievement
- Parents set SMART goals with students
- Teachers share strategies for practicing skils at home
- Inviting parents in to school has added benefit of improving safety and culture


